|

### 2017-09-25 manuscript.Rmd
finish addressing Dirk edit list.

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly,
using a layered filesystem which minimizes disk footprint and download time,
and are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.

@@

All Rocker images are based on the Debian Linux distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the `debian:stable`
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker project
can be largely divided into two stacks which address different needs,
defined by which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on `debian` stable releases.  These stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

1. These Dockerfiles require occasional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
<!--4. `testing` and `unstable` do not receive the attention to security updates from the Debian security team \citep{debian_security}.-->
<!-- DE: Are you sure about that point 4 ? -->
<!-- CB: I dunno, just quoting from https://www.debian.org/releases/testing/ -->
<!-- DE: Which says nothing about _unstable_ not getting security updates, which is implied here-->

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time (but will continue to reconstruct the same environment in
the future after newer R versions are released) should explicitly include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

@@

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
code-named `jessie`, while `3.4.0`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occasionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should never install R
packages, using `apt`, since this will install a second, older version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

    wget -qO- https://get.docker.com/ | sh 
    sudo docker run -d -p 80:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio

Many academic and commercial cloud providers make it
possible to execute such code snippets when an instance is launched, without
ever needing to `ssh` into the machine. The user may log into the server merely
by pasting its IP address or DNS name into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

@@

Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
associated with the most recent date for which that image was current.  This ensures that
a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
writing) will continue to advance on the Docker-hub image until the next R release.  Software
installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
using `bioclite()` utility will also install the version appropriate to the version of R found on the system
(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
or adopt a more comprehensive solution such as `packrat` \citep{packrat}; however, `packrat` should not be necessary
for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
A more general discussion of the use of Docker for computational reproducibility can be found in \citep{Boettiger2015}.

@@

Sharing these Dockerfiles can also facilitate the emergence 
of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
image emerged from the input of a number of Rocker users all adding common
geospatial libraries and packages on top of the existing Rocker images. 
This coalescence helped create a more fine-tuned image with broad support 
for a wide range of commonly used data formats and libraries.  Other community
images are developed and maintained independently of the Rocker project,
such as the `popgen` image of population-genetics-oriented software developed 
by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
being used as base Docker images in the NSF sponsored Whole Tale project for
reproducible computing \citep{wholetale}.

>>>>

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly
using a layered filesystem which minimizes disk footprint and download time,
and are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.

@@

All Rocker images are based on the Debian Linux distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the `debian:stable`
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker project
can be largely divided into two stacks which address different needs,
defined by which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on Debian stable releases.  These stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

1. These Dockerfiles require occasional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the then-contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time (but will continue to reconstruct the same environment in
the future after newer R versions are released) should explicitly include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

@@

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
code-named `jessie`, while `3.4.0`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occasionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should avoid installing R
packages using `apt` without careful consideration, as this will install a second, 
probably different version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

    wget -qO- https://get.docker.com/ | sh 
    sudo docker run -d -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio

Many academic and commercial cloud providers make it
possible to execute such code snippets when an instance is launched, without
ever needing to `ssh` into the machine. The user may log into the server merely
by pasting its IP address or DNS name (followed by the chosen port, RStudio's default `:8787` in the above example)
into a browser and entering the
appropriate password.  This provides the user with a familiar,
interactive environment running on a remote machine while requiring
a minimum of expertise.

@@

Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
associated with the most recent date for which that image was current.  This ensures that
a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
writing) will continue to advance on the Docker-hub image until the next R release.  Software
installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
using `bioclite()` utility will also install the version appropriate to the version of R found on the system
(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
or adopt an alternative approach such as `packrat` \citep{packrat}. A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \citep{Boettiger2015}. 
<!--
however, `packrat` should not be necessary
for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
-->

@@

Sharing these Dockerfiles can also facilitate the emergence 
of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
image emerged from the input of a number of Rocker users all adding common
geospatial libraries and packages on top of the existing Rocker images. 
This coalescence helped create a more fine-tuned image with broad support 
for a wide range of commonly used data formats and libraries.  Other community
images are developed and maintained independently of the Rocker project,
such as the `popgen` image of population-genetics-oriented software developed 
by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
being used as base Docker images in the NSF sponsored Whole Tale project for
reproducible computing \citep{wholetale}, and are heavily used by the `rhub`
project in automated package testing \citep{rhub}.
